---
title: "Chi Squared Testing for Project"
date: "`r Sys.Date()`"
output: pdf_document
---

# Read in Data

```{r}

# Read in electric car data
e_df <- read.csv('../data/cardata_electric_clean.csv')

# Read in non-electric car data
ne_df <- read.csv('../data/cardata_nonelectric_clean.csv')

(nrow(e_df))
(nrow(ne_df))
head(e_df)
head(ne_df)
```
```{r}
# Create color palettes
Blues <- colorRampPalette(c("#0A146B", "#A9A3DA")) 
Purples <- colorRampPalette(c("#3E1370", "#BDA3DA")) 
GrBuPuPi <- c("#095826", "#0E7032", "#10913F", "#55A472", "#8CBF9E", "#8CBFB8", 
              "#63B7AC", "#2D9A8B", "#137568", "#094E45", "#0B3C5C", "#17547C", 
              "#2671A4", "#3C8CC1", "#72B1DB", "#96C3E1", "#B0CDE1", "#B0B3E1", 
              "#858ACD", "#4F55AB", "#1923B3", "#0E1468", "#3C1075", "#5821A1", 
              "#6B27C4", "#9455E5", "#A278D8", "#A990CA", "#ADA0BF", "#C1A5CB", 
              "#B887CA", "#A35CBD", "#762594")
```

# EDA

To formulate our hypotheses, we first perform EDA on the dataset

```{r}
#Vehicle.Manufacturer.Name
#CO2..g.mi.
library(ggplot2)
library(tidyr)

# Drop NAs for the emissions column
ne_df <- ne_df %>% drop_na(CO2..g.mi.)

ne_df$emission_cat[ne_df$CO2..g.mi. < 250] <- "low"
ne_df$emission_cat[ne_df$CO2..g.mi. >= 250 & ne_df$CO2..g.mi. < 500] <- "medium"
ne_df$emission_cat[ne_df$CO2..g.mi. >= 500] <- "high"

head(ne_df)
```
```{r}
# Make bar plot of transmission type and CO2 emissions
ggplot(data=ne_df, aes(x=Vehicle.Manufacturer.Name,y=CO2..g.mi.))+geom_boxplot()+ scale_x_discrete(guide = guide_axis(angle = 45))
```

```{r}
# make barplot of emissions categories

ggplot(ne_df, aes(x=Vehicle.Manufacturer.Name, fill=emission_cat)) + geom_bar(position="dodge")+ scale_x_discrete(guide = guide_axis(angle = 45)) + scale_fill_manual(values = GrBuPuPi[c(5, 14, 20)])

```


We definitely see some car manufacturers have higher average fuel emissions. For instance, Lamborghini, Bentley, and Rolls-Royce have higher emissions likely due to them being luxury brands. Honda and Mitsubishi, on the other hand, are more affordable brands and have lower average emissions.

```{r}
# Make bar plot of transmission type and CO2 emissions
ggplot(data=ne_df, aes(x=Tested.Transmission.Type,y=CO2..g.mi., fill=CO2..g.mi.))+geom_boxplot()+ scale_x_discrete(guide = guide_axis(angle = 45)) + scale_fill_manual(values = GrBuPuPi[c(1,5, 10, 16, 22, 27, 29, 30)])
```

```{r}
ggplot(ne_df, aes(x=Tested.Transmission.Type, fill=emission_cat)) + geom_bar(position="dodge")+ scale_x_discrete(guide = guide_axis(angle = 45)) + scale_fill_manual(values = GrBuPuPi[c(5, 14, 20)])
```
Interestingly, it appears like the high emissions cars are mostly within the automatic and semi-automatic categories. Manual tends to have mostly low and medium with very few high emissions cars. 


Drive system vs Emission category. Looks like 2 wheel drive front does not have high emissions compared to the 4 wheel drive and all wheel drive. Perhaps fuel emissions increase the more the wheel drive increases.

```{r}
ggplot(ne_df, aes(x=Drive.System.Description, fill=emission_cat)) + geom_bar(position="dodge")+ scale_x_discrete(guide = guide_axis(angle = 45)) + scale_fill_manual(values = GrBuPuPi[c(5, 14, 20)])
```
Next, check observe the transmission type vs the drive system type

```{r}
ggplot(ne_df, aes(fill=Drive.System.Description, x=Tested.Transmission.Type)) + geom_bar(position="dodge")+ scale_x_discrete(guide = guide_axis(angle = 45)) + scale_fill_manual(values = GrBuPuPi[c(1, 7, 15, 25, 32)])
```


# Define Hypotheses

**Question 1:** Is there a relationship between car brand and fuel emission level? 

Null hypothesis: Car Brand and fuel emissions level are independent. The fuel emissions level does not depend on the car brand

Alternative Hypothesis: Car Brand and fuel emissions level are dependent. The fuel emissions level does depend on the car brand                                                                                          


**Question 2:** Is there a relationship between transmission type and fuel emission level? 

Null hypothesis: Transmission type and fuel emissions level are independent. The fuel emissions level does not depend on the transmission type

Alternative Hypothesis: Transmission type and fuel emissions level are dependent. The fuel emissions level does depend on the Transmission type      


**Question 3:** Is there a relationship between transmission type and the drive system type? 

Null hypothesis: Transmission type and drive system type are independent. The transmission type does not depend on the drive system type

Alternative Hypothesis: Transmission type and drive system type are dependent. The transmission type does depend on the drive system type      


**Question 4:** Is there a relationship between drive system and the fuel emission level type? 

Null hypothesis: Drive system and fuel emissions level are independent. The fuel emissions level does not depend on the drive system

Alternative Hypothesis: Drive System and fuel emissions level are dependent. The fuel emissions level does depend on the drive system 



# Hypothesis Testing

```{r}
head(ne_df)
```

## Question 1

**Question 1:** Is there a relationship between car brand and fuel emission level? 

Null hypothesis: Car Brand and fuel emissions level are independent. The fuel emissions level does not depend on the car brand

Alternative Hypothesis: Car Brand and fuel emissions level are dependent. The fuel emissions level does depend on the car brand  
 
```{r}
# Make contingency table
cont <- table(ne_df$Vehicle.Manufacturer.Name, ne_df$emission_cat)
cont
```
```{r}
chisq.test(cont)
```

Repeat test with the Yates correction
```{r}
chisq.test(cont, correct = TRUE)
```

Since the Yates correction was not enough, we can switch over to Fisher's Exact Test
```{r}
fisher.test(cont, simulate.p.value=TRUE)
```

## Question 2

**Question 2:** Is there a relationship between transmission type and fuel emission level? 

Null hypothesis: Transmission type and fuel emissions level are independent. The fuel emissions level does not depend on the transmission type

Alternative Hypothesis: Transmission type and fuel emissions level are dependent. The fuel emissions level does depend on the Transmission type  

```{r}
# Make contingency table
cont <- table(ne_df$Tested.Transmission.Type, ne_df$emission_cat)
cont
```

```{r}
chisq.test(cont)
```

Repeat test with the Yates correction
```{r}
chisq.test(cont, correct = TRUE)
```

Since the Yates correction was not enough, we can switch over to Fisher's Exact Test
```{r}
fisher.test(cont, simulate.p.value=TRUE)
```

## Question 3

**Question 3:** Is there a relationship between transmission type and the drive system type? 

Null hypothesis: Transmission type and drive system type are independent. The transmission type does not depend on the drive system type

Alternative Hypothesis: Transmission type and drive system type are dependent. The transmission type does depend on the drive system type     

```{r}
# Make contingency table
cont <- table(ne_df$Tested.Transmission.Type, ne_df$Drive.System.Description)
cont
```

```{r}
chisq.test(cont)
```

Repeat test with the Yates correction
```{r}
chisq.test(cont, correct = TRUE)
```

Since the Yates correction was not enough, we can switch over to Fisher's Exact Test
```{r}
fisher.test(cont, simulate.p.value=TRUE)
```

**Question 4:** Is there a relationship between drive system and the fuel emission level type? 

Null hypothesis: Drive system and fuel emissions level are independent. The fuel emissions level does not depend on the drive system

Alternative Hypothesis: Drive System and fuel emissions level are dependent. The fuel emissions level does depend on the drive system 

```{r}
# Make contingency table
cont <- table(ne_df$Drive.System.Description, ne_df$emission_cat)
cont
```

Repeat test with the Yates correction
```{r}
chisq.test(cont, correct = TRUE)
```

Since the Yates correction was not enough, we can switch over to Fisher's Exact Test
```{r}
fisher.test(cont, simulate.p.value=TRUE)
```

